---
title: "RMS Titanic"
author: "Olivia Smith"
date: 3/3/2025
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
glimpse(df_titanic)
```

**Observations**:

Class, age, sex, Survived, n

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
- Are there any differences?
  - Yes, the Wikipedia page says that 2,224 souls were aboard while the data set says that there are 2201 souls.
- If yes, what might account for those differences?
  - With so many people aboard the ship, perhaps different passenger rosters had differing and erroneous data. The Wikipedia article says that the 2,224 souls aboard
  is an 'estimate', so perhaps they were trying to account for stowaways or people not legally a passenger.
  - ...

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
survivors <- df_titanic %>% filter(Survived == "Yes")
survivors %>% 
  ggplot(aes(fill = Sex, x = Class, y = n)) +
  geom_col(position = "Stack")
```

**Observations**:

- I noticed that while in 1st, 2nd, and 3rd class Females were the sex that survived the most, while within the Crew the vast majority of survivors were men. I also
noticed that the Crew was the highest demographic of 'class' that survived, and 2nd class was the class with the least survivors. 
# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
# survivors <- 
#   df_titanic %>% 
#   group_by(Sex, Class)  %>% 
#   mutate(
#     Total = sum(n),
#     Prop = n / Total
#   ) %>% 
#   filter(Survived == "Yes")


# survivors %>% 
df_prop %>%
  ggplot(aes(fill = Sex, x = Class, y = Prop)) +
  geom_col(position = "Dodge")
```

**Observations**:

- Write your observations here.
  -I noticed that most of the survivors are from first and second class.
  -According to this graph, the only deaths were in third class and the crew.
  - Most fatalities are in third class. 
  
- Is there anything *fishy* going on in your plot?
- I noticed that the graph says that all 1st and 2nd class passengers survived, while that is not true. 

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
survivors <- df_prop %>% filter(Survived == "Yes")

survivors %>% 
  ggplot(aes(fill = Sex, x = Class, y = Prop)) +
  geom_col(position = "dodge") +
  facet_grid(Age ~ .)
```

**Observations**:

- In all classes except Crew, children had a higher chance of survival. Perhaps this is because there were no children employed by the Titanic, or so few that they became a statistical anomaly. However, children in the third class had about the same chance of survival as the adults.Among the adults, women had a much higher chance of survival while the gender divide was fairly equal among the children.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - In my graph above, I was questioning why the ratio of women to men survivors was so high despite women being given precedence on the lifeboats. From this graph I can see that this is probably because a large proportion of the survivors are children, and children were allowed on the boats regardless of gender. This would have somewhat evened out the gender ratio of the survivors.  

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
