---
title: "Antibiotics"
author: "Olivia Smith"
date: 3/9/2025
output:
  github_document
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective. The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*. A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()

```

# Visualization
<!-- -------------------------------------------------- -->

### __q1__ Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`. These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long. You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
df_antibiotics %>% 
  pivot_longer(cols = penicillin:neomycin, names_to = "antibiotic") %>% 
  mutate(bacteria = fct_reorder(bacteria, value)) %>% 
  ggplot(aes(x = bacteria, y = log(value), fill = antibiotic)) +
  geom_col(position = "dodge" , aes(color = gram)) +
  geom_hline(yintercept = 0.1, color = 'black', linetype = "longdash") +
  coord_flip()

```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}
df_antibiotics %>% 
  pivot_longer(cols = penicillin:neomycin, names_to = "antibiotic") %>% 
  mutate(bacteria = fct_reorder(bacteria, value)) %>% 
  ggplot(aes(x = bacteria, y = log(value))) +
  geom_hline(yintercept = 0.1, color = 'black', linetype = "longdash") +
  geom_col(position = "dodge", aes(fill = gram)) +
  facet_grid(antibiotic ~ .) + 
  coord_flip()
  
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}
df_antibiotics %>% 
  pivot_longer(cols = penicillin:neomycin, names_to = "antibiotic") %>% 
  mutate(bacteria = fct_reorder(bacteria, value)) %>% 
  filter(antibiotic == "streptomycin") %>% 
  ggplot(aes(x = value, fill = bacteria)) +
  geom_bar(position = "stack", stat = "count")
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE
df_antibiotics %>% 
  pivot_longer(cols = penicillin:neomycin, names_to = "antibiotic") %>% 
  ggplot(aes(x = antibiotic, y = log(value), color = gram), position = 'jitter') +
  geom_point() +
  geom_hline(yintercept = 0.1, color = 'black', linetype = "longdash")
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
df_antibiotics %>% 
  pivot_longer(cols = penicillin:neomycin, names_to = "antibiotic") %>% 
  mutate(bacteria = fct_reorder(bacteria, value)) %>% 
  ggplot(aes(x = bacteria, y = log(value), color = antibiotic)) +
  geom_point() + 
  facet_grid(gram ~ .) +
  geom_hline(yintercept = 0.1)
  
```

### __q2__ Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions. Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below. You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions. **Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

*Observations*
- What is your response to the question above?
  - Penicillin is effective against some bacteria, but requires a higher dosage. As MIC must be lower for effective human treatment, at most a .1 concentration, this implies that penicillin is ineffective for many bacteria strains including all the gram negative bacterias.  Streptomycin is also effective against many bacteria, but requires a very low dose. It is a pretty even mix of negative and positive gram. Similarly, Neomycin is effective with a low dose and is pretty evently split between negative and positive gram. 
  
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - Visual 2
- Why?
  - Visual 2 offered comparison for all the antibiotics/bacteria as well as their gram status. This allowed for a more in-depth analysis than the other figures. Figure 1 also offered these metrics, but in a more difficult to read way. 

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

*Observati ons*
- What is your response to the question above?
  - Perhaps it was renamed because Diplococcus pneumoniae is of positive gram like the streptococcus bacteria, so it is related to it.
  
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - 1
  
- Why?
  - 1 helped a lot because the bacteria was grouped together on the y-axis and thus easier to compare to one another. It also offered all the distinguishing factors being considered, so I could see how Diplococcus pneumoniae could be related to streptococcus. 


# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
